A Theory for Data Migration of Information Systems

### 2022-09-23 2022 Migration/articleMigration.tex
introduced rules and working on a second example for the to-be system.

\section{Terminology}
\label{sct:Terminology}
	To migrate data from one information system to another,
	we must define ``information system''.
	In that definition we will single out the dataset as the focal point of a data migration.
	For this reason, we define datasets first.

\subsection{Datasets}
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
	The purpose of an information system is to make data meaningful to users.
	Data migration is all about preserving that meaning.
	So this section introduces the notion of information system
	as a system that consist of a dataset and a set constraints that users must keep satisfied.
	These constraints reflect the meaning that is shared by all users.

	Users have their own tasks and responsibilities
	and may work from different locations and on different moments.
	This collective use serves a purpose which we loosely call ``the business''.
	As a consequence, the data in an information system changes continually.
	To preserve meaning, users are trying to maintain semantic constraints on the data,
	amidst of all changes that are going on around them.
	So, let us first define the constituent notions role and constraint before defining information systems proper.

@@

	A \define{constraint} is a restriction on a dataset, to be enforced by a role.
	The purpose of constraints is to ensure the semantics of a dataset, i.e. the preservation of meaning as data changes.
	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
	We say that a constraint \define{is satisfied} if its violation set is empty.
	The set of constraints is called $\rules$.

	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
	Since $\triples$ changes continually in an information system, the violations are changing along.

	We can now define the notion of information system.
	Let $\roles$ be a set of roles,
	let $\rules$ be a set of constraints, 
	let $\maintain$ be a surjective relation $\roles\times\rules$,
	and let $\dataset$ be a dataset.
\begin{definition}[information system]
	\label{def:information system}
	\item An information system $\infsys$ is a tuple $\la\roles,\rules,\maintain,\dataset\ra$\\
	with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
	in which $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{\triples}{u}$ empty.
\end{definition}
	If a constraint $u$ has violations, i.e. $\viol{\triples}{u}$ is not empty, there is a role whose job it is to remove those violations.
	The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
	Which means that all constraints are being kept satisfied in the collective effort of users, even if the data changes.
	In another section we will give some options on how users can keep rules satisfied.

\subsection{Example}
	Having defined an information system in mathematical terms, let us give an example.
	For this purpose we use a subset of the language Ampersand~\cite{Joosten-JLAMP2018}
	because it makes the examples more appealing to read.
	Let us first define a dataset of eight triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
	This code snippet introduces a relation with name \verb#takes#,
	source concept \verb#Student#, and
	target concept \verb#Course#.
	It also introduces three triples:
\[\begin{array}{l}
	\triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
	\triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
	\triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
\end{array}\]
	The informal meaning of this relation is that it states which students are taking which courses.
	So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.
	As stated before, this informal meaning is interesting only for the sake of illustration.

@@

	Now let us define an information system.
	To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
	We define a constraint that states that a student can enroll for any module that is part of a course that student takes.
	In logic, this constraint is written as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]
	In Ampersand, which is a syntactically sugared form of relation algebra,
	we give that constraint a name and state a role to maintains it:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
	The signature of this rule is $\pair{\text{\tt Student}}{\text{\tt Module}}$, which is derived by the Ampersand compiler.
	Ampersand also derives the violation set as:
\[\begin{array}{l}
	\viol{\triples}{\text{\tt EnrollRule}}\\
	\hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{array}\]
	Notice that the current population yields an empty set of violations,
	so the rule is satisfied.
	Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
	we have now defined a complete information system (albeit small).

@@

	We use two datasets: $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ and $\dataset'=\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra$.
	Before doing so, let us first define the disjoint union of two information systems.
\begin{definition}[disjoint union of datasets]
	\begin{eqnarray}
		\dataset\sqcup\dataset'&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\\
		X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
		X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
		X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
		X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}
	\end{eqnarray}
\end{definition}
Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.

@@

\begin{definition}[disjoint union of information systems]
	\begin{eqnarray}
		\infsys\sqcup\infsys'&=&\la\roles\uplus\roles',\rules\uplus^?\rules',\maintain\uplus^2\maintain',\dataset\sqcup\dataset'\ra
	\end{eqnarray}
\end{definition}

>>>>

\section{Terminology}
\label{sct:Terminology}
	To migrate data from one information system to another,
	we must define ``information system''.
	In that definition we will single out the dataset as the focal point of a data migration.
	For this reason, we define datasets first.
	We need rules to capture the semantics, so we need to define rules as well.
	After that, we define information systems.

\subsection{Datasets}
\label{sct:Datasets}
	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.

@@

\subsection{Rules}
\label{sct:Rules}
	A \define{rule} is used to restrict the populations of a dataset.
\begin{definition}[rule]
	\label{def:rule}
	\item A rule $u$ is a tuple $\la p_u,A,B,\id{viol}_u\ra$
	in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
	$\pair{A}{B}$ is the signature of $u$, and
	$\id{viol}_u:\powerset{\triples}\rightarrow\powerset{\atoms\times\atoms}$ is a function that gives the violations (which are pairs) of this rule,
	in which
\begin{eqnarray}
	\pair{a}{b}\in\viol{u}{\triples}&\Rightarrow&a\in A\wedge b\in B\\
	p_u&\Leftrightarrow&\viol{u}{\triples}=\emptyset
\end{eqnarray}
\end{definition}
	The function $\id{viol}_u$ computes the violations of $u$ for any set of triples $\triples$ in the dataset,
	which allows us to check at any moment in time whether the current population is free of violations.
	We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
	Since $\triples$ changes continually in an information system, the violations are changing along.
	As a result, rules that are violated at one moment may be satisfied at another moment.

	The set of triples changes due to events observed by the dataset.
	These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
	Every rule is assigned to a role, so that role must ensure that the rule returns to a satisfied state after some event has caused the set of triples to change.
	In other words: If anyone edits in the dataset and breaks a rule, someone has to ensure that the data is changed so that satisfies the rule.

\subsection{Information Systems}
\label{sct:Information Systems}
	The purpose of an information system is to make data meaningful to users.
	Data migration is all about preserving that meaning.
	This section introduces the notion of information system
	as a system that consist of a dataset and a set of rules that constrain that data,
	together with a set of roles that must keep all rules satisfied.
	These rules reflect the meaning (semantics) that is shared among users.

	Users have their own tasks and responsibilities
	and may work from different locations and on different moments.
	This collective use serves a purpose which we loosely call ``the business''.
	As a consequence, the data in an information system changes continually.
	To preserve meaning, users are trying to maintain semantic constraints on the data,
	amidst of all changes that are going on around them.
	So, let us first define the constituent notions role and rule before defining information systems proper.

@@

	We can now define the notion of information system.
	Let $\roles$ be a set of roles,
	let $\rules$ be a set of rules, 
	let $\maintain$ be a surjective relation $\roles\times\rules$,
	and let $\dataset$ be a dataset.
\begin{definition}[information system]
	\label{def:information system}
	\item An information system $\infsys$ is a tuple $\la\roles,\rules,\maintain,\dataset\ra$\\
	with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
	in which $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
\end{definition}
	If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
	The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
	Which means that all constraints are being kept satisfied in the collective effort of users, even if the data changes.
	In another section we will give some options on how users can keep rules satisfied.

\subsection{Example}
\label{As-is IS}
	Having defined an information system in mathematical terms, let us give an example.
	For this purpose we use a subset of the language Ampersand~\cite{Joosten-JLAMP2018}
	because it makes the examples more appealing to read.
	Let us first define a dataset of eight triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
	This code snippet introduces a relation with name \verb#takes#,
	source concept \verb#Student#, and
	target concept \verb#Course#.
	It also introduces three triples:
\[\begin{array}{l}
	\triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
	\triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
	\triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
\end{array}\]
	The informal meaning of this relation is that it states which students are taking which courses.
	So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.
	As stated before, this informal meaning is interesting only for the sake of illustration.

@@

	Now let us define an information system.
	To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
	We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
	Its constraint, $p_{\tt Enroll}$ is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]
	In Ampersand, which is a syntactically sugared form of relation algebra,
	we give rule a name and declare a role to maintains it:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
	The Ampersand compiler derives the signature of every rule.
	This rule has signature $\pair{\text{\tt Student}}{\text{\tt Module}}$.
	The Ampersand compiler also derives the violation set as:
\[\begin{array}{l}
	\viol{\text{\tt EnrollRule}}{\triples}\\
	\hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{array}\]
	Notice that the current population yields an empty set of violations,
	so the rule is satisfied.
	Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
	we have now defined a small information system.

@@

	We use two datasets: $\dataset=$ and $\dataset'=$.
	Before doing so, let us first define the disjoint union of two information systems.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
	\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra
\end{eqnarray}
\begin{eqnarray}
	\omit\rlap{Some text that I want to appear here...}\\
	&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\\
		X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
		X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
		X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
		X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
\end{eqnarray}
\end{definition}
Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.

\subsection{Example}
	Let us proceed to specify a to-be information system, by way of example.
	We will use the example in section~\ref{As-is IS} as the as-is information system.
	The new information system keeps all three relations, but it constrains one relation, \verb-isPartOf-, to be univalent.
	This means that every \verb-Module- is part of at most one \verb-Course-.
	It restricts the relation \verb-isPartOf-,
	so the migration might yield violations in the to-be information system that were not in the as-is information system.
\begin{verbatim}
RELATION takes[Student*Course]
RELATION isPartOf[Module*Course] [UNI]
RELATION isEnrolledFor[Student*Module]
RELATION course[ExamReg*Course] [UNI]
RELATION student[ExamReg*Student] [UNI]
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
RULE ExamRule1: student~;course |- takes
RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
	Part of the change is that exam registrations (concept \verb-ExamReg-) will be involved.
	In an exam registration, a student registers for the examination of a course.
	The to-be information information system contains two extra rules: ExamRule1 and ExamRule2.
	The first one says that an exam registration requires that a student actually takes the course.
	In logic, ExamRule1 is written as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
	e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
\end{array}\]
	Another requirement, ExamRule2, is that the student is enrolled for every module that is part of the course.
	In logic its constraint, $p_{\tt ExamRule2}$ is written as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
	e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
\end{array}\]
	Ampersand also derives the following violation sets:
\[\begin{array}{l}
	\viol{\text{\tt ExamRule1}}{\triples}\\
	\hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
	\viol{\text{\tt ExamRule2}}{\triples}\\
	\hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
\end{array}\]

@@

\begin{definition}[]
	\begin{eqnarray}
		\infsys\sqcup\infsys'&=&\la\roles',\rules',\maintain',\dataset\sqcup\dataset'\ra
	\end{eqnarray}
\end{definition}
